---
title: "SurvToGit"
author: "Adrian Reuschel"
date: "2024-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

LOAD PACKAGES
```{r}
library(dplyr)
# install.packages("dplyr")
library(survival)
library(MASS)
# update.packages(ask = FALSE, dependencies = TRUE)
```

DATA Source:
https://www.sciencedirect.com/science/article/pii/S2352340920301876#appsec2

LOAD DATA
```{r}
#NEW DATASET 
god <- read.csv("/Users/abbyreuschel/Desktop/pls/god.csv")
```


DATA CLEANING
```{r}
# #trimming for columns's I want to explore
beetle <- god %>%
  select(CellID, Year, MPB, Latitude, Longitude, ColdTolerance, BP1, BP2, BP1red, BP2red, BP1man, BP2man)

#renaming col's
new_column_names <- c("Group", "Monitoring_Period", "Delta", "Site_Lat", "Site_Long", "Cold_Tolerance", "BP1_Infected_Prev","BP2_Infected_Prev", "BP1_NoTreatment_Prev", "BP2_NoTreatment_Prev", "BP1_Treatment_Prev", "BP2_Treatment_Prev")


beetle <- setNames(beetle, new_column_names)
```



```{r}
beetle_tte <- beetle %>%
  select(Group, Monitoring_Period, Delta, Site_Lat, Site_Long, Cold_Tolerance, BP1_Infected_Prev, BP1_Treatment_Prev, BP2_Infected_Prev, BP2_Treatment_Prev)

#deleting incomplete groups
beetle_tte <- beetle_tte %>%
  group_by(Group) %>%
  filter(all(c(2006:2018) %in% Monitoring_Period)) %>%
  ungroup()

#making deltas after first 1 to 0
beetle_tte <- beetle_tte %>%
  group_by(Group) %>%
  mutate(Delta = ifelse(row_number() > which.max(Delta == 1), 0, Delta)) %>%
  ungroup()

#adding Time col
beetle_tte$Time <- beetle_tte$Monitoring_Period

observation_count <- table(beetle_tte$Monitoring_Period)
print(observation_count)
```

Next Steps:
- Figure out how to use coordinates to delineate "sites"
    - how big is area covered?
- Learn about treatment types in this study
- Read into the area and problems associated with the pine beetle
- Formulate research question to answer using methods of original paper

(1) ABOUT THE DATA
Region: Saskatchewan, Western province of Canada
Location: Cypress Hills
- inhabited by multiple indigenous groups
- data collected by Saskatchewan Ministry of Environment Forest Service
- groups are determined with a 50m radius
- 2006 outbreak of pine beetles led to this data monitoring



K-Means Clustering, Spatial Dependence  

```{r}
library(ggplot2)
#install.packages("cluster")
library(cluster)
```

```{r}
# subset1 <- beetle_tte %>%
#   group_by(Group) %>%
#   arrange(Monitoring_Period) %>%
#   slice(1) %>%
#   ungroup()

locations <- beetle_tte[, c("Site_Lat", "Site_Long")]
set.seed(2023)
k <- 4
result <- kmeans(locations, centers = k, nstart = 10)


ggplot(beetle_tte, aes(x = Site_Long, y = Site_Lat, color = factor(result$cluster))) +
  geom_point() +
  labs(title = "Clusters of Sites")
```

```{r}
cluster_counts <- table(result$cluster)
cluster_counts
```

Kmeans result:
4 clusters is the best, each have 5500-6200 observations.

```{r}
#assign sites to each group
beetle_tte$Site <- result$cluster
```


Research Question:
Are the control methods employed on areas of pine beetle infested trees effective in slowing the spread of the pine beetles to a 1 or 2 degree radius?

- Learn about treatment types in this study
 # cut and burn
- Read into the area and problems associated with the pine beetle
 # beetle is about the size of a grain of rice
 # biggest threat to pine trees in Canada
 # an outbreak in british columbia killed over 50%  of pine trees
 # https://www.saskatchewan.ca/residents/environment-public-health-and-safety/saskatchewan-state-of-the-environment-2023/mountain-pine-beetle-detection-and-prevention#:~:text=MPB%20is%20the%20most%20significant,in%20Saskatchewan%20and%20across%20Canada 



Recreating Methods of Original Paper
------------------------------------
Models
------
1. Full Model
- classic coxph model, with monitoring period and site

Survival Model
```{r}
# Full model
library(survival)
surv_mod <- Surv(beetle_tte$Time, beetle_tte$Delta)
```

Risk Indicator Function
```{r}
# Risk indicator function
# Y = 1 (subject at ith site in kth monitoring period at risk) or 0 (not at risk)

beetle2 <- beetle_tte[order(beetle_tte$Site, beetle_tte$Monitoring_Period), ]

beetle2 <- beetle2 %>%
  group_by(Group) %>%
  mutate(
    Y_ki = ifelse(row_number() <= which.max(Delta == 1), 1, 0)
  ) %>%
  ungroup()


# Checking indicator function
risk_event_table <- beetle2 %>%
  group_by(Monitoring_Period) %>%
  summarize(
    TotalY = sum(Y_ki),
    TotalEvents = sum(Delta)
  )
risk_event_table
```

Checking Risk Function
```{r}
# Check
# 
# events_2007 <- beetle2 %>%
#   filter(Monitoring_Period == 2007, Delta ==1) %>%
#   select(Group, Delta, Monitoring_Period, Y_ki)
# events_2007
# 
# #look at 884, 1145, 1271 and check that 2006 they had events- all look correct!!
# #look at 913, 2809, 18893 and check that 2007 they are not considered at risk- all look correct!!
# 
# group_884_obs <- beetle2 %>%
#   filter(Group == 884) %>%
#   select(Group, Delta, Monitoring_Period, Y_ki)
# group_884_obs
# 
group_18893_obs <- beetle2 %>%
  filter(Group == 18893) %>%
  select(Group, Delta, Monitoring_Period, Y_ki)
group_18893_obs
# 
# # Rest of 2007 events to check: 874*, 908*,1165*, 1169*, 1515,1650,1918,2144,2929,18489
# # * = checked and good!
# 
# #its because the events are totaled and are accumulating!! it actually looks correct-- fix the table code if needed.

```

Coxph options
- A strata term identifies a stratified Cox model; separate baseline hazard functions are fit for each strata.

- A time-transform term allows variables to vary in time. In this case the tt argument will be a function or a list of functions (if there are more than one tt() term in the model) giving the appropriate transform.

Full Model, COXPH... what goes wrong?
```{r}
# Full model
# beetle2$Time <- as.numeric(beetle2$Monitoring_Period)
# 
# surv_mod <- Surv(beetle2$Time, beetle2$Delta)
# 
# fullmod_data <- data.frame(
#   SurvObj = surv_mod,
#   Y_ki = beetle2$Y_ki,
#   covariate1 = beetle2$BP1_Infected_Prev,
#   covariate2 = beetle2$BP1_Treatment_Prev,
#   covariate3 = beetle2$BP2_Infected_Prev,
#   covariate4 = beetle2$BP2_Treatment_Prev, 
#   site = as.factor(beetle2$Site),
#   Time = beetle2$Time
# )
# 
# cox_model <- coxph(SurvObj ~ Y_ki + 
#                                 tt(covariate1) + 
#                                 tt(covariate2) + 
#                                 tt(covariate3) +
#                                 tt(covariate4) +
#                                 site +
#                                 tt(covariate1): tt(covariate2) +
#                                 tt(covariate3): tt(covariate4) +
#                                 tt(covariate1): site +
#                                 tt(covariate2): site +
#                                 tt(covariate3): site +
#                                 tt(covariate4): site
#                    ,data = fullmod_data)
# 
# # Graphing and summaries
# summary(cox_model)
#Convergence issues in coxph model.. paper says glm() will fix these computational inefficiencies
```
-convergence issues
-non-significant parameters


Considering GLM w/ logloglink

"glm function - RDocumentation." Home - RDocumentation, www.rdocumentation.org/packages/stats/versions/3.6.2/topics/glm. Accessed 4 Dec. 2023.
"Modelling a binary outcome when census interval varies." Cross Validated, stats.stackexchange.com/questions/148699/modelling-a-binary-outcome-when-census-interval-varies/148728#148728. Accessed 4 Dec. 2023.

```{r}
link_df <- data.frame(
  SurvObj = surv_mod,
  Y_ki = beetle2$Y_ki,
  covariate1 = beetle2$BP1_Infected_Prev,
  covariate2 = beetle2$BP1_Treatment_Prev,
  covariate3 = beetle2$BP2_Infected_Prev,
  covariate4 = beetle2$BP2_Treatment_Prev, 
  site = as.factor(beetle2$Site),
  Time = beetle2$Time,
  event = as.numeric(beetle2$Delta)
)

#baseline mod
link_base <- glm(event ~ Time, family = binomial(link = 'cloglog'), data = link_df)
summary(link_base)

#full mod
link_full <- glm(event ~ offset(Y_ki) +
                        Time +
                        covariate1:Time + 
                        covariate2:Time + 
                        covariate3:Time + 
                        covariate4:Time + 
                        (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) + 
                        site +
                        covariate1:site + 
                        covariate2:site + 
                        covariate3:site + 
                        covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df)

summary(link_full)
```

GLM Model Selection
```{r}
#using AIC
library(MASS)
aic_full <- stepAIC(link_full, direction = "backward")
#same result
full_final <- glm(event ~ Time + site + 
                    Time:covariate1 + Time:covariate2 + Time:covariate3 + Time:covariate4 + 
                    (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +  
                    covariate1:site + covariate2:site + covariate3:site + covariate4:site + 
                    offset(Y_ki),
    family = binomial(link = 'cloglog'),
    data = link_df)
summary(full_final)
```
Model selected after AIC:

event ~ Time + site + Time:covariate1 + Time:covariate2 + 
    Time:covariate3 + Time:covariate4 + (covariate1:Time):(covariate2:Time) + 
    (covariate3:Time):(covariate4:Time) + covariate1:site + covariate2:site + 
    covariate3:site + covariate4:site + offset(Y_ki)

AIC = 12965


Checking assumptions for model (1)
- goodness of fit
- Check residuals
```{r}
anova(link_base, full_final, test ="Chisq")
#test shows that model significantly improves 
link_base$aic
full_final$aic
link_base$aic - full_final$aic
#lowered aic by 3549.653
```

```{r}
#deviance residuals
dev_residuals1 <- residuals(full_final, type = 'deviance')

plot(link_df$Time, dev_residuals1,
     type = 'b', pch = 10, 
     xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals for GLM Model 1")
```

```{r}

#checking for outliers because of deviance plot
plot(link_df$Time, dev_residuals1,
     type = 'b',
     xlab = 'time', ylab = 'dev residuals',
     main = "Mod1: Residuals with Outliers")

points(link_df$Time[outliers1], dev_residuals1[outliers1], pch = 18, col = 'red')

outliers1 <- which(dev_residuals1 > 3 | dev_residuals1 < -3)
outlier_sites1 <- link_df$site[outliers1]
print(outlier_sites1)
#get a table
site_counts <- table(link_df$site[outliers1])
site_counts
```
-only one outlier in site 2 (furthest west)

What happens when I remove outliers??
```{r}
#full mod, no outliers
link_df_no_outliers <- link_df[-outliers1, ]


link_outliers <- glm(event ~ offset(Y_ki) +
                        Time +
                        covariate1:Time + 
                        covariate2:Time + 
                        covariate3:Time + 
                        covariate4:Time + 
                        (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) + 
                        site +
                        covariate1:site + 
                        covariate2:site + 
                        covariate3:site + 
                        covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df_no_outliers)

summary(link_outliers)
```
-estimates are much better

Model selection without outliers
```{r}
aic_outliers <- stepAIC(link_outliers, direction = "backward")
#
full_final_no_outliers <- glm(event ~ Time + site + 
                                Time:covariate1 + Time:covariate3 + Time:covariate4 + 
                                (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +  
                                covariate1:site + covariate2:site + covariate3:site + covariate4:site + 
                                offset(Y_ki),
                              family = binomial(link = "cloglog"), data = link_df_no_outliers)
summary(full_final_no_outliers)

```
Best model from outlier-free mod:

event ~ Time + site + Time:covariate1 + Time:covariate3 + 
    Time:covariate4 + covariate1:covariate2 + covariate3:covariate4 + 
    covariate1:site + covariate2:site + covariate3:site + covariate4:site + 
    offset(Y_ki)
    
AIC = 8447.7
- the first step of the model selection process removed the same variable
- only 2 units less than before model selection

```{r}
#comparing AIC's
#base model
link_base$aic
#full model
link_full$aic
#final with outliers
full_final$aic
#final without outliers
full_final_no_outliers$aic
```
```{r}
#Model 1 interpretations:
summary(full_final)
```
- in this model, the predictor that increases log-odds of individual at ith site in kth monitoring period getting infested by bark beetles is the individual being at site 2.
- investigate site 2 to draw more conclusions about this... did site 2 have most events?
- other significant predictors of log odds were:
    site3                  4.701e-01  1.116e-01   4.213 2.52e-05 ***
    site4                  2.953e-01  1.175e-01   2.514 0.011938 *  
    Time:covariate1        7.866e-04  8.131e-05   9.674  < 2e-16 ***
    Time:covariate3        5.037e-04  7.607e-05   6.622 3.54e-11 ***
    Time:covariate4        4.104e-04  1.412e-04   2.906 0.003662 ** 
    Time:(covariate1:covariate2) -2.160e-04  4.013e-05  -5.382 7.36e-08 ***  This says the treatment is somewhat mitigating prob
    Time:(covariate3:covariate4) -1.284e-04  1.695e-05  -7.578 3.51e-14 ***  mitigating, but less sig than 1 deg radius
    site2:covariate1      -1.134e+00  1.791e-01  -6.334 2.39e-10 *** impact of cov 1 decreasing over time
    site3:covariate1      -6.462e-01  2.112e-01  -3.060 0.002214 **  impact of cov 1 decreasing over time
    site2:covariate2       9.058e-01  1.771e-01   5.116 3.12e-07 *** impact of cov 2 increasing over time
    site3:covariate2       7.980e-01  2.592e-01   3.079 0.002078 **  impact of cov 2 increasing over time
    site2:covariate3      -5.287e-01  1.605e-01  -3.294 0.000989 *** impact of cov 3 decreasing over time
for each unit increase in infected trees in a previous year (1deg rad), the effect of treatment trees during the prev year decreases the log odds of the event happening by -4.243e-01


Likelihood Ratio for Hypothesis
```{r}
#model without treatment covariates
lrt_red1 <- glm(event ~ Time + site + 
                    Time:covariate1 + Time:covariate3 + 
                    covariate1:site + covariate3:site +
                    offset(Y_ki),
    family = binomial(link = 'cloglog'),
    data = link_df)
#model with treatment covariates
lrt_full1 <- glm(event ~ Time + site + 
                    Time:covariate1 + Time:covariate2 + Time:covariate3 + Time:covariate4 + 
                    (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +  
                    covariate1:site + covariate2:site + covariate3:site + covariate4:site + 
                    offset(Y_ki),
    family = binomial(link = 'cloglog'),
    data = link_df)
#chisq test
lrt1 <- anova(lrt_red1, lrt_full1, test = "Chisq")
lrt1
```
- LRT shows including treatment predictors improves the model performance





Reduced Models
---------------
2. Reduced Mod 1
- constant Beta

COXPH reduced mod 1... what goes wrong? doesn't converge
- homogeneous beta, heterogeneous Z vector



GLM Reduced mod 1
```{r}
link_df <- data.frame(
  SurvObj = surv_mod,
  Y_ki = beetle2$Y_ki,
  covariate1 = beetle2$BP1_Infected_Prev,
  covariate2 = beetle2$BP1_Treatment_Prev,
  covariate3 = beetle2$BP2_Infected_Prev,
  covariate4 = beetle2$BP2_Treatment_Prev, 
  site = as.factor(beetle2$Site),
  Time = beetle2$Time,
  event = beetle2$Delta
)

#reduced mod 1
link_red1 <- glm(event ~ Time + 
                        offset(Y_ki) + 
                        covariate1 + 
                        covariate2 + 
                        covariate3 + 
                        covariate4 + 
                        covariate1:covariate2 + 
                        covariate3:covariate4 + 
                        site +
                        covariate1:site + 
                        covariate2:site + 
                        covariate3:site + 
                        covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df)
summary(link_red1)
```
Model Selection GLM reduced 1
```{r}
#using AIC backward
aic_red1 <- stepAIC(link_red1, direction = "backward")
```
Reduced model after selection
```{r}
link_red1final <- glm(event ~ Time + offset(Y_ki) + covariate1 + covariate2 + covariate3 + 
    covariate4 + covariate1:covariate2 + covariate3:covariate4 + 
    site + covariate1:site + covariate2:site + covariate3:site + 
    covariate4:site,
    family = "binomial"(link= 'cloglog'), data = link_df)
summary(link_red1final)
```

Checking COXph assumptions for model (2)
- proportionality
- LRT for goodness of fit
- Check residuals

```{r}
#deviance residuals
dev_residuals2 <- residuals(link_red1final, type = 'deviance')

plot(link_df$Time, dev_residuals2,
     type = 'b', pch = 10, 
     xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals for GLM Model 2")
```
```{r}
#these outliers are the same
plot(link_df$Time, dev_residuals2,
     type = 'b',
     xlab = 'time', ylab = 'dev residuals',
     main = "Mod1: Residuals with Outliers")

points(link_df$Time[outliers1], dev_residuals1[outliers1], pch = 18, col = 'red')

outliers2 <- which(dev_residuals2 > 3 | dev_residuals2 < -3)
outlier_sites2 <- link_df$site[outliers2]
print(outlier_sites2)
#get a table
site_counts2 <- table(link_df$site[outliers2])
site_counts2
```

Reduced mod 1 no outliers
```{r}
#red mod1, no outliers
link_df_no_outliers2 <- link_df[-outliers2, ]


link_outliers2 <- glm(event ~ offset(Y_ki) +
                        Time +
                        covariate1 + 
                        covariate2 + 
                        covariate3 + 
                        covariate4 + 
                        covariate1:covariate2 + 
                        covariate3:covariate4 + 
                        site +
                        covariate1:site + 
                        covariate2:site + 
                        covariate3:site + 
                        covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df_no_outliers2)

summary(link_outliers2)
```
Model selection without outliers
```{r}
aic_outliers2 <- stepAIC(link_outliers2, direction= 'backward')
```
red mod1 w/o outliers final:
```{r}
link_red1final_no_outliers <- glm(event ~ Time + covariate1 + covariate2 + covariate3 + covariate4 + 
    site + covariate1:covariate2 + covariate3:covariate4 + covariate1:site + 
    covariate3:site + covariate4:site + offset(Y_ki),
    family = "binomial"(link = "cloglog"), data = link_df_no_outliers2)
summary(link_red1final_no_outliers)
```

Checking AIC's:
```{r}
#base
link_base$aic
#full final
full_final$aic
#full no outliers
full_final_no_outliers$aic
#red mod1 final
link_red1final$aic
#red mod1 final no outliers
link_red1final_no_outliers$aic
```

- reduced model barely lower AIC, still lower though

Mod red1 interpretations
```{r}
summary(link_red1final)
```
The parameter that most impacted log-odds of beetle infestation in this model was covariate 1 which is the num of infested tree cells in 1 cell radius. covariate1..    for unit increase in num infested trees, the log odds of an individual experiencing the event increases by 1.560958   (< 2e-16 ***).
Other significant parameters:
    covariate3             1.015004   0.152727   6.646 3.01e-11 *** increase in infected... increase in log odds
    covariate4             0.832855   0.282492   2.948 0.003196 **  increase of treatment (2deg) also increase in log odds
    site2                  1.146065   0.101592  11.281  < 2e-16 *** site 2 most diff from site 1
    site3                  0.471928   0.111661   4.226 2.37e-05 ***
    site4                  0.297341   0.117517   2.530 0.011400 *  
    covariate1:covariate2 -0.434844   0.080750  -5.385 7.24e-08 *** cov 2 decreases cov 1 impact (evidence for hypothesis) 
    covariate3:covariate4 -0.258518   0.034099  -7.581 3.42e-14 *** same thing here just lesser degree
    covariate1:site2      -1.110107   0.194359  -5.712 1.12e-08 ***
    covariate1:site3      -0.622361   0.223265  -2.788 0.005311 ** 
    covariate2:site2       0.770139   0.387065   1.990 0.046625 *  
    covariate3:site2      -0.530832   0.160221  -3.313 0.000923 ***

Likelihood Ratio for Hypothesis
```{r}
#model without treatment covariates
lrt_red2 <- glm(event ~ Time + offset(Y_ki) + covariate1 + covariate3 + 
    site + covariate1:site + covariate3:site,
    family = binomial(link = 'cloglog'),
    data = link_df)
#model with treatment covariates
lrt_full2 <- glm(event ~ Time + offset(Y_ki) + covariate1 + covariate2 + covariate3 + 
    covariate4 + covariate1:covariate2 + covariate3:covariate4 + 
    site + covariate1:site + covariate2:site + covariate3:site + 
    covariate4:site,
    family = "binomial"(link= 'cloglog'), data = link_df)
#LRT
lrt2 <- anova(lrt_red2, lrt_full2, test = "Chisq")
lrt2
```
- LRT test on reduced model 2 shows addition of treatment parameters improves model 



3. Reduced Mod 2
- stratify surv function on site and monitoring period
- groupedsurv package?
- try hybrid model to lower AIC
```{r}
#groupedsurv package
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("qvalue")
install.packages("groupedSurv")
library(groupedSurv)
```
https://www.bioconductor.org/packages/release/bioc/html/qvalue.html

```{r}
#estimating baseline survival on diff monitoring times
mod3surv <- alphaEstFam(beetle2$Monitoring_Period, beetle2$Delta)
mod3surv
```
 https://cran.r-project.org/web/packages/groupedSurv/groupedSurv.pdf
 


GLM reduced mod2 (Hybrid)
- look at outputs... what looks time dependent and what doesn't

```{r}
summary(link_red1final)
summary(full_final)

#time dependent covariates affect odds less (smaller estimates, so make homogeneous)
#include time interaction for interaction between treatment and infected covariates

link_hybrid <- glm(event ~ Time + offset(Y_ki) + 
                     covariate1 + covariate2 + covariate3 + covariate4 + 
                     (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +
                      site + 
                     covariate1:site + covariate2:site + covariate3:site + covariate4:site,
    family = "binomial"(link= 'cloglog'), data = link_df)
summary(link_hybrid)
```

Reduced Mod2, model selection
```{r}
aic_hybrid <- stepAIC(link_hybrid, direction = 'backward')
```
```{r}
#deviance residuals
#these also same
dev_residuals4 <- residuals(link_hybrid, type = 'deviance')

plot(link_df$Time, dev_residuals4,
     type = 'b', pch = 10, 
     xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals for GLM Model Hybrid")
```



Reduced mod 2 no outliers
```{r}
#red mod2, no outliers
link_df_no_outliers2 <- link_df[-outliers2, ]


link_hybrid2 <- glm(event ~ Time + offset(Y_ki) + 
                     covariate1 + covariate2 + covariate3 + covariate4 + 
                     (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +
                      site + 
                     covariate1:site + covariate2:site + covariate3:site + covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df_no_outliers2)

summary(link_hybrid2)
```

```{r}
aic_hybrid_no_outliers <- stepAIC(link_hybrid2, direction = "backward")
```

Best model:
event ~ Time + covariate1 + covariate2 + covariate3 + covariate4 + 
    site + covariate1:site + covariate3:site + covariate4:site + 
    Time:covariate1:covariate2 + Time:covariate3:covariate4 + 
    offset(Y_ki)

AIC = 8445.36

```{r}
hybrid_final_no_outliers <- glm(event ~ Time + 
                                  covariate1 + covariate2 + covariate3 + covariate4 + 
                                  site + 
                                  covariate1:site + covariate3:site + covariate4:site + 
                                  Time:covariate1:covariate2 + Time:covariate3:covariate4 + 
                                  offset(Y_ki),
                                family = "binomial"(link="cloglog"), data = link_df_no_outliers2)
```

AIC's
```{r}
#base
link_base$aic
#full final
full_final$aic
#full no outliers
full_final_no_outliers$aic
#red mod1 final
link_red1final$aic
#red mod1 final no outliers
link_red1final_no_outliers$aic
#hybrid final
link_hybrid$aic
#hybrid final no outliers
hybrid_final_no_outliers$aic
```

Checking COXph assumptions for model (3)
- proportionality
- LRT for goodness of fit
- Check residuals

```{r}
#I'm doing model without outliers for this visual
#final model gave best AIC

#deviance residuals
dev_residuals3 <- residuals(hybrid_final_no_outliers, type = 'deviance')

plot(link_df_no_outliers$Time, dev_residuals3,
     type = 'b', pch = 10, 
     xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals for GLM Model 3")

outliers5 <- which(dev_residuals3 > 3 | dev_residuals3 < -3)
outlier_sites5 <- link_df$site[outliers5]
print(outlier_sites5)
#get a table
site_counts5 <- table(link_df$site[outliers5])
site_counts5
```
- only 2 at site two 
- i have 118 more outliers in this set


```{r}
#removing more outliers
link_df_no_outliers2 <- link_df_no_outliers[-outliers5, ]


hybrid_final2 <- glm(event ~ Time + offset(Y_ki) + 
                     covariate1 + covariate2 + covariate3 + covariate4 + 
                     (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +
                      site + 
                     covariate1:site + covariate2:site + covariate3:site + covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df_no_outliers2)

summary(hybrid_final2)
aic_hybrid2 <- stepAIC(hybrid_final2, direction = "backward")
```

Actual Final Final Hybrid Fit without outliers
```{r}
hybrid_final2_no_outliers <- glm(event ~ Time + 
                                   covariate1 + covariate2 + covariate3 + covariate4 + 
                                    site + 
                                  covariate1:site + covariate3:site + 
                                   Time:covariate1:covariate2 + Time:covariate3:covariate4 + 
                                   offset(Y_ki),
                                 family = binomial(link = 'cloglog'), data = link_df_no_outliers2
)
summary(hybrid_final2_no_outliers)
```


Model Selection
AIC's
```{r}
#base model, no predictors
link_base$aic
#full model, before removing outliers
full_final$aic
#full model, after removing outliers
full_final_no_outliers$aic
#model 2, before removing outliers
link_red1final$aic
#model 2, after removing outliers
link_red1final_no_outliers$aic
#hybrid model, before removing outliers
link_hybrid$aic
#hybrid model, after removing outliers
hybrid_final3_no_outliers$aic
```
Outlier Justification

```{r}
library(dplyr)
outlier_investigation <- beetle2[outliers1, ]
head(outlier_investigation)
infections_prev1 <- table(outlier_investigation$BP1_Infected_Prev)
infections_prev1
infections_prev2 <- table(outlier_investigation$BP2_Infected_Prev)
infections_prev2
events_outliers <- table(outlier_investigation$Delta)
events_outliers
#they all have events, but no previous infections to 1 or 2 degree radius
#this indicates a "random"ish appearance of beetles
#My research question is more interested in treatment methods, and these outliers didn't even receive treatment bc there were no recorded infections
```
```{r}
outlier_investigation2 <- beetle2[outliers5, ]
head(outlier_investigation2)
at_risk <- table(outlier_investigation2$Y_ki)
at_risk
#these except for 3 all observed the event already
```

Model Assumptions:
```{r}
dev_residuals6 <- residuals(hybrid_final3_no_outliers, type = 'deviance')

plot(link_df_no_outliers3$Time, dev_residuals6,
     type = 'b', pch = 10, 
     xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals for Best Model")

outliersfinal <- which(dev_residuals6 > 3 | dev_residuals6 < -3)
outlier_sites6 <- link_df$site[outliersfinal]
print(outlier_sites6)
#get a table
site_counts5 <- table(link_df$site[outliersfinal])
site_counts5
```
```{r}
outlier_investigation3 <- beetle2[outliersfinal, ]
head(outlier_investigation3)
at_risk2 <- table(outlier_investigation3$Y_ki)
at_risk2
#observed event already

#removing more outliers
link_df_no_outliers3 <- link_df_no_outliers2[-outliersfinal, ]


hybrid_final3 <- glm(event ~ Time + offset(Y_ki) + 
                     covariate1 + covariate2 + covariate3 + covariate4 + 
                     (covariate1:Time):(covariate2:Time) + 
                        (covariate3:Time):(covariate4:Time) +
                      site + 
                     covariate1:site + covariate2:site + covariate3:site + covariate4:site
                        ,
                 family = binomial(link = 'cloglog'), data = link_df_no_outliers3)

summary(hybrid_final3)
aic_hybrid2 <- stepAIC(hybrid_final3, direction = "backward")
```
```{r}
hybrid_final3_no_outliers <- glm(event ~ Time + 
                                   covariate1 + covariate2 + covariate3 + covariate4 + 
                                    site + 
                                   covariate1:site + covariate2:site + covariate3:site + 
                                    covariate1:covariate2 + covariate3:covariate4 + 
                                  offset(Y_ki),
                                 family = 'binomial'(link= 'cloglog'), data = link_df_no_outliers3)
summary(hybrid_final3_no_outliers)
```

```{r}
#model assumptions
dev_residuals7 <- residuals(hybrid_final3_no_outliers, type = 'deviance')

plot(link_df_no_outliers3$Time, dev_residuals7,
     type = 'b', pch = 10, 
     xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals for Best Model")

outliersfinal2 <- which(dev_residuals7 > 3 | dev_residuals7 < -3)
outlier_sites7 <- link_df$site[outliersfinal2]
print(outlier_sites7)
#get a table
site_counts6 <- table(link_df$site[outliersfinal2])
site_counts6
#only 25 outliers in this deviance plot
```

```{r}
#checking for outliers because of deviance plot
plot(link_df_no_outliers3$Time, dev_residuals7,
     type = 'b',
     xlab = 'time', ylab = 'dev residuals',
     main = "Mod3: Residuals with Outliers")

points(link_df_no_outliers3$Time[outliersfinal2], dev_residuals7[outliersfinal2], pch = 18, col = 'red')

```

Likelihood Ratio for Hypothesis

Fang, Qixiang, and Rans van de Schoot. "Intro to Discrete-Time Survival Analysis in R." Rans van de Schoot, www.rensvandeschoot.com/tutorials/discrete-time-survival. Accessed 4 Dec. 2023.
```{r}
#final model without treatment covariates
lrt_red1 <- glm(event ~ Time + 
                                   covariate1 + covariate3 + 
                                    site + 
                                   covariate1:site + covariate3:site + 
                                  offset(Y_ki),
                                 family = 'binomial'(link= 'cloglog'), data = link_df_no_outliers3)
# final model with treatment covariates
lrt_full1 <- glm(event ~ Time + 
                                   covariate1 + covariate2 + covariate3 + covariate4 + 
                                    site + 
                                   covariate1:site + covariate2:site + covariate3:site + 
                                    Time:covariate1:covariate2 + Time:covariate3:covariate4 + 
                                  offset(Y_ki),
                                 family = 'binomial'(link= 'cloglog'), data = link_df_no_outliers3)
#LRT
lrt1 <- anova(lrt_red1, lrt_full1, test = "Chisq")
lrt1

```

Model Interpretation : Best Model
```{r}
summary(hybrid_final3_no_outliers)
```

Interpretations
----------------
Coefficients:
                             Estimate Std. Error z value Pr(>|z|)    
(Intercept)                 1.262e+02  2.264e+01   5.574 2.49e-08 ***
Time                       -6.743e-02  1.126e-02  -5.988 2.12e-09 *** 
covariate1                  3.030e+00  2.870e-01  10.559  < 2e-16 *** infections in prev year do impact current year 1 deg
covariate2                 -8.683e-02  5.568e-01  -0.156 0.876087     NOT significant... treatment not impacting enough
covariate3                  1.324e+00  2.211e-01   5.988 2.13e-09 *** infections do impact but less than 1 deg radius
covariate4                  1.186e+00  1.236e-01   9.600  < 2e-16 *** significant but increase still... same conclusion as ''
site2                       4.204e+00  3.871e-01  10.861  < 2e-16 *** most significant impact on log odds...least outliers too
site3                       9.987e-01  4.433e-01   2.253 0.024264 *   3 is next to 2, so evidence for spatial impact
covariate1:site2           -2.559e+00  2.970e-01  -8.616  < 2e-16 ***
covariate1:site3           -1.386e+00  3.460e-01  -4.005 6.19e-05 *** 
covariate1:site4           -7.614e-01  3.959e-01  -1.923 0.054432 .  
covariate2:site2            1.202e+00  5.364e-01   2.240 0.025075 *  
covariate2:site3            1.176e+00  5.872e-01   2.003 0.045198 *  
covariate3:site2           -7.906e-01  2.240e-01  -3.530 0.000416 ***
Time:covariate1:covariate2 -2.803e-04  4.561e-05  -6.146 7.97e-10 *** as time, num infected and num treated increase,
Time:covariate3:covariate4 -1.739e-04  1.906e-05  -9.120  < 2e-16 *** the log-odds decrease. this suggests treatment slows
                                                                      spread of infection, but not fast enough to be sig alone
